A Transformer Model for Retrosynthesis

### 2019-04-04 papers/retrosynthesis/article.tex
minor changes

\begin{abstract}
We describe a Transformer model for a retrosynthetic reaction prediction task. 
The model is trained end-to-end on 45\,033 experimental reaction examples extracted 
from USA patents. Our final model can successfully predict the reactants set for 
42.7\% of cases on the external test set. During the training procedure, we applied 
different learning rate schedules and snapshot learning. These techniques can prevent 
overfitting and thus can be a reason to get rid of internal validation dataset that 
is advantageous for deep models with millions of parameters. We investigate 
the influence of the temperature while decoding the model output probabilities and find 
that at T=1.3 the final model can improve the accuracy up to 1-2\%. 
We also discuss the synergy of applying together the retrosynthetic model to predict 
reagents and a direct reaction model trained on the same data which makes a prognosis 
for the product base on reactants.
\end{abstract}

\keywords{Reaction outcomes prediction \and Computer retrosynthesis \and Character-based models \and Transformer}

@@

The analogy between machine translation and retrosynthesis is evident: each target molecule has 
its predecessors from which it can be synthesized as every meaningful sentence one can translate 
from source language to target one. If all parts of a reaction are written in SMILES notation, 
then our source and target sentence are composed of valid SMILES tokens as words. 
The main goal of the work is to build a model which could for a given target molecule for our example\footnote{
This reaction is in the test set and it was correctly predicted  by our model. 
Probability landscapes in the following figures are drawn from this example.} COC(=O)c1cccc(-c2nc3cccnc3[nH]2)c1 in fig. \ref{fig:example-reaction}
correctly predict the set of reactants. Namely, it should predict Nc1cccnc1N.COC(=O)c1cccc(C(=O)O)c1 in this case.

@@

If we apply the early stopping technique, we should select the model around 200 epoch. Effectiveness of such a model marked $T1_1$ in table \ref{tbl:results} resulted in TOP-1 37.9\% on the test set. If we choose the last one model obtained at 1\,000 epoch, then the model $T1_2$ gives us better values -- 39.8\%. In this case, we did not see any need of the validation dataset and keeping in mind that our model has almost 2 millions of parameters we decided to combine training and validation sets and train our next models on both data, e.g., without validation. The model $T2$ was trained on all data and with the same learning rate schedule as $T1$. The results obtained when applying T2 to the test set are better than for T1 set namely 41.8\% against 39.8\%.

@@

It worth to notice that TOP-5 accuracy reaches almost 70\%. That means the model can correctly predict reactants but sometimes scoring is wrong and TOP-1 is much less. We tried to improve TOP-1 scoring with two different approaches: internal confidence estimation and forward reaction prediction.

@@

\subsection{Forward reaction prediction scoring}

We built a forward reaction prediction model based on the same data and with the same parameters as we did for retrosynthesis. The final model correctly predicted 84\% of test data. Then we embedded the model in the following algorithm. Once the retrosynthesis model returned possible sets of reactants, we applied the forward model for each item in this set, and the forward model made a prognosis for a product. Then we compared the predicted and real outcomes, and if they both were the same, this reactant set was used as the solution to a particular problem. The example of this procedure is illustrated below for a target molecule:

\noindent\textbf{Cc1cnc(N2CCN(C(=O)c3ccc(N4CCCS4(=O)=O)nc3C)CC2)c(C)c1}

The first beam returns  C[S+](C)(C)=O.Cc1nc(Br)ccc1C(=O)N1CCN(c2ccc(N3CCCS3(=O)=O)c(C)n2)CC1.
And estimation of a product Cc1nc(Br)ccc1C(=O)N1CCN(c2ccc(N3CCCS3(=O)=O)c(C)n2)CC1 differs from the target. Continue analyzing other beams.

The second beam contains C[Mg+].Cc1nc(Br)ccc1C(=O)N1CCN(c2ccc(N3CCCS3(=O)=O)c(C)n2)CC1 and the forward model predicts   Cc1ccc(C(=O)N2CCN(c3ccc(N4CCCS4(=O)=O)c(C)n3)CC2)c(C)n1 as a product and again it is not we expect.

Next beam with Cc1nc(Br)ccc1C(=O)N1CCN(c2ccc(N3CCCS3(=O)=O)c(C)n2)CC1.Cc1cnc(N2CCNCC2)c(C)c1  gives a product: Cc1cnc(N2CCN(c3ccc(C(=O)N4CCN(c5ccc(N6CCCS6(=O)=O)c(C)n5)CC4)c(C)n3)CC2)c(C)c1, which is again wrong.

At last the fourth beam predicts valid reactants COC(=O)c1ccc(N2CCCS2(=O)=O)nc1C.Cc1cnc(N2CCNCC2)c(C)c1 and the forward model  assembles the right product from them

\textbf{Cc1cnc(N2CCN(C(=O)c3ccc(N4CCCS4(=O)=O)nc3C)CC2)c(C)c1}.

This schema strongly depends on the quality of the forward model.

@@

>>>>

\begin{abstract}
We describe a Transformer model for a retrosynthetic reaction prediction task. 
The model is trained end-to-end on 45\,033 experimental reaction examples extracted 
from USA patents. Our final model can successfully predict the reactants set for 
42.7\% of cases on the external test set. During the training procedure, we applied 
different learning rate schedules and snapshot learning. These techniques can prevent 
overfitting and thus can be a reason to get rid of internal validation dataset that 
is advantageous for deep models with millions of parameters. We thoroughly investigated different 
approaches to train Transformer models and found snapshot learning with averaging weights on 
learning rates minima works best.  While decoding the model output probabilities there is a strong 
 influence of the temperature that improves at T=1.3 the accuracy of models up to 1-2\%. 
\end{abstract}

\keywords{Retrosynthesis prediction \and Computer Aided synthesis Planning \and Character-based models \and Transformer}

@@

The analogy between machine translation and retrosynthesis is evident: each target molecule has 
its predecessors from which it can be synthesized as every meaningful sentence one can translate 
from source language to target one. If all parts of a reaction are written in SMILES notation, 
then our source and target sentence are composed of valid SMILES tokens as words. 
The main goal of the work is to build a model which could for a given target molecule for our example\footnote{
This reaction is in the test set and it was correctly predicted  by our model.} COC(=O)c1cccc(-c2nc3cccnc3[nH]2)c1 in fig. \ref{fig:example-reaction}
correctly predict the set of reactants. Namely, it should predict Nc1cccnc1N.COC(=O)c1cccc(C(=O)O)c1 in this case.

@@

If we apply the early stopping technique, we should select the model around 200 epoch. Effectiveness of such a model marked T1$_1$ in table \ref{tbl:results} resulted in TOP-1 37.9\% on the test set. If we choose the last one model obtained at 1\,000 epoch, then the model T1$_2$ gives us better value -- 39.8\%. In this case, we did not see any need of the validation dataset and keeping in mind that our model has almost 2 millions of parameters we decided to combine training and validation sets and train our next models on both data, e.g., without validation. The model T2 was trained on all data and with the same learning rate schedule as T1. The results obtained when applying T2 to the test set are better than for T1 set namely 41.8\% vs. 39.8\%.

@@

It worth to notice that TOP-5 accuracy reaches almost 70\%. That means the model can correctly predict reactants but sometimes scoring is wrong and TOP-1 is much less. We tried to improve TOP-1 scoring with internal confidence estimation.

@@

@@

Our model can correctly predict reactant set in TOP-5 with accuracy 69.8\%. Internal confidence estimation cannot guarantee a correct solution, so different scoring methods should be developed. One of the promising ways is to use a forward reaction prediction model to estimate whether it is possible to assemble a target molecule from reactants proposed. The scoring model should have excellent characteristics and probably it is possible to apply the same cycling learning rate and snapshot averaging to build it.
